Identifying optimal baselines for stable isotope studies of stream food webs

### 2022-12-13 docs/ms.Rmd
new abstract and updating text

**Type of manuscript**: Article

@@

Stable isotope analysis (SIA) is used in aquatic ecosystems to reconstruct food webs and deduce trophic interactions. Because land use strongly influence d15N values in consumers, suitable isotopic baselines must be established for meaningful cross-site comparisons of food webs. 
Our objective was to identify an invertebrate taxon or Functional Feeding Group (FFG) useful as a baseline for d15N values in riverine food webs. The ideal baseline should meet four criteria: 1) be easy to collect with wide geographic distribution, 2) have low within-site d15N variation among individuals, 3) have d15N values that are correlated with land use and background d15N values, and 4) produce no systemic change in calculated consumer trophic positions with % land use (i.e. be independent of system-specific d15N variability).  We investigated these criteria in invertebrate communities in 16 sites along a gradient from mountain to prairie streams which also represent a gradient in land use.

@@

We sampled sixteen 2nd-5th order stream sites distributed along the Rocky Mountain—Great Plains ecotone. Sites were selected to represent an environmental gradient based on elevation and land use that resulted in a wide variation of d15N values at the base of local food webs. Sites spanned the longitudinal (i.e., upstream-downstream) gradient of the three primary tributaries to the North Platte River in Wyoming, USA (Figure 1). The Sweetwater, Medicine Bow and Laramie rivers are snowmelt dominated with peak flows occurring in the spring and base flows occurring in mid-summer. Their watersheds transition from forested land at high elevations to agriculturally dominated land with some urbanization at low elevations.

### Longitudinal gradient and land use index variable

Instead of using percent natural land cover as a variable to compare d15N values among potential baselines (Kristensen et al., 2016), we used a proxy variable for a site’s location along the longitudinal stream gradient developed by Maitland (2020). The proxy variable combines seven metrics of environmental conditions at a study site into an index of longitudinal stream position using the Axis 1 site score from a principal component analysis (PCA). Input variables to the PCA were elevation (m a.s.l.), stream slope (km/km), distance to North Platte River (km), Strahler stream order, mean August water temperature (℃), upstream drainage area (km2), and stream width (m). This longitudinal gradient index (PC1) explained 78% of site-level variation among the 16 study sites and described changes in environmental conditions from upstream, high-elevation, cold-water streams to downstream, low-elevation, warm-water streams. Upstream sites were more likely to be surrounded by natural land cover (e.g. forest, grassland, shrubland, and wetland land types; NLCD Database 2011), as evidenced by a strong negative correlation between PC1 and percent natural landcover at a site (Pearson's correlation, r = -0.69, t15 = -3.66, P < 0.001; Maitland, 2020). Maitland (2020) also found that PC1 (and thus the degree of natural landcover) was positively correlated with d15N values of basal resources and primary consumers (Linear Regression - Basal Resources: F1,373 = 121.4, P < 0.001; Primary Consumers: F¬¬1,14 = 4.7, P = 0.050), consistent with previous research (Cabana and Rasmussen 1996, Lake et al. 2001, Kristensen et al. 2016). Thus, we used PC1 as a measure of local variation in d15N values for all future analyses. 
Sample collection and preparation

Samples were collected during summer 2016. Basal resources (i.e., primary uptake compartments) were sampled once in June, July, and August to account for fine-scale temporal variation in d15N (Jardine et al., 2014).  Seston was collected by filtering three replicates of up to 10 L of water onto pre-combusted (550 °C, 4 hours) filters (Whatman GF/F, 47 mm O) in the field using a modified portable drill pump (Kelso & Baker, 2016). Biofilm was collected by scraping five cobble-sized rocks within the reach.  Macroinvertebrates were collected during the same sampling events during the summer of 2016. A D-framed kick net was used in a variety of habitats (i.e. riffle, pools, macrophytes, and river margins) to collect a representative sample of the macroinvertebrate assemblage.  Macroinvertebrates were kept in filtered stream water for 24 hours to allow for gut clearance, then frozen. Muscle tissue samples were obtained from fish collected during the August sampling event using a backpack electrofishing unit (Smith-Root, Vancouver, WA).  Brown Trout Salmo trutta (Linnaeus, 1758), Creek Chub Semotilus atromaculatus (Person, Year), White Sucker Catostomus commersonii (Lacépède, 1803), were used because these species were found at greater than 50% of the sites.  To minimize the number of fish euthanized, a muscle plug using a 5-mm biopsy punch was taken from fish greater than 300 mm in length. The remaining fish used for SIA were euthanized with a lethal dose of MS-222, placed on ice and then frozen upon returning to the lab. Those fish not selected for SIA were returned to the stream.

Macroinvertebrates were identified to family level using an invertebrate guide (Merritt et al. 2008). We classified macroinvertebrates into FFGs using a trait-based method. We obtained invertebrate trophic trait data from the USA Freshwater Biological Traits Database (Vieira et al. 2006), and then assigned trait affinity scores from 0 (absent) to 3 (strong affinity) using “fuzzy-coding” (Chevene et al. 1994). Because larvae and adults have different feeding behaviors, Elmidae were separated by life stage during identification.

@@

\begin{equation}
\delta^{n} = (\frac{R_{sample}}{R_{standard}}+1) * 100 (\#eq:ratio)
\end{equation}

where dN is either N15 or c13, and R is the ratio of the heavy to lighter isotope. Analytical error (i.e., 1 SD of lab standard) of sample runs was estimated at 0.07‰ (d13C) and 0.18‰ (d15N) for the invertebrates, and 0.04‰ (d13C) and 0.10‰ (d15N) for the fish.

### Baseline selection

Suitability of various macroinvertebrate taxa as baselines were based on the four criteria proposed by Kristensen et al. (2016).  The baseline taxon must: (1) have a wide geographic distribution, (2) have low with-in site variation of d15N values, (3) have d15N values correlated with environmental influences of d15N, (4) give trophic positions estimates of other consumers free of environment influences of d15N.  Although all analyses were performed for all macroinvertebrate taxa and FFGs, selection occurred progressively from each criterion. For example, if a macroinvertebrate taxon met the first criterion it would stay in the list of potential baselines and be evaluated for the second criterion, but if it did not meet the first criterion then it would be removed from the list.

@@

The geographic distribution was expressed as the percentage of the 16 sites that contained a given macroinvertebrate taxon. We used a cut off of 75% of sites in the selection process as a reasonable value for identifying a widely distributed taxon.  Although 75% was chosen arbitrarily, Kristenson et al. (2016) included two taxa that would fall below this mark, and the best baseline identified by Kristenson et al. (2016) was found at 78.9% of sites. This makes our cutoff a more conservative measure of distribution while retaining the distribution levels of the best baselines in other studies.

@@

We calculated the coefficient of variation (CV) of d15N values for all macroinvertebrate taxa and FFGs. A low coefficient of variation for d15N indicates that a baseline candidate has a non-omnivorous diet. To calculate CV, we divided the standard deviation of d15N signatures by the mean d15N signature of each baseline candidate at each site.  We then found the mean and 95% confidence intervals for each baseline candidate across sites.

We use one-way analysis of variance (ANOVA) to test for differences in CV values among baselines. Separate analyses were done on individual taxa and FFGs.  The CV values were square root or log transformed to ensure normality.  We used the Tukey HSD post hoc tests for pairwise-comparison using an alpha of 0.05.

@@

For all macroinvertebrate taxa and FFGs, we used least squares linear regression to assess relationships between the baseline’s d15N position at sites along the longitudinal stream gradient.  We wished to compare our baselines’ relationships to the longitudinal gradient with those of the basal resources. So, we did separate least squares linear regressions for two of the basal resources at our sites and compared R2 values between the baselines and the resources at our sites.  The R2 value of our basal resources represent the minimum level of variation that our alternate baselines should explain in order to be used as alternatives to the traditional method (e.g. all basal resources across a season).   
<!-- Baselines with significant ($P \le 0.05$) relationships with the longitudinal gradient and that had R2 values equal to or greater than the basal resources were considered suitable baselines. -->

>>>>

**Type of manuscript**: Article in *Freshwater Biology*

@@

Optimal baselines for stable isotope-based studies of trophic structure in Rocky Mountain--Great Plains streams

SIA is a helpful tool for quantifying trophic structure in stream ecosystems.

Predictable changes in the ratio of heavy to light N isotopes (15N/14N; δ15N) in tissues between food resources and consumers can be used to assess their relative trophic position in a food web.

However, cross-site comparisons of trophic structure cannot be done without suitable isotopic baselines because nitrogen isotope ratios (15N/14N, d15N) in aquatic biota can vary substantially across space due to anthropomorphic N loading. Further, consensus about optimal screening criteria for baseline selection is lacking.

Here, we assess the applicability and utility of a recently proposed screening procedure for identifying suitable baseline organisms in Danish stream systems by applying it to streams spanning the Rocky Mountain Great Plains ecotone in North America. Individual taxa and aggregated groups representing similar feeding modes were hierarchically screen based on 4 criteria: 1) easy to collect with a wide distribution, 2) low within-site d15N variation, 3) d15N values reflect system-specific d15N variability, 4) produce consumer trophic positions independent of system-specific d15N variability.

Stable isotope analysis is used in aquatic ecosystems to reconstruct food webs and deduce trophic interactions. Because land use strongly influence d15N values in consumers, suitable isotopic baselines must be established for meaningful cross-site comparisons of food webs. Our objective was to identify an invertebrate taxon or Functional Feeding Group (FFG) useful as a baseline for d15N values in riverine food webs. The ideal baseline should meet four criteria: 1) be easy to collect with wide geographic distribution, 2) have low within-site d15N variation among individuals, 3) have d15N values that are correlated with land use and background d15N values, and 4) produce no systemic change in calculated consumer trophic positions with % land use (i.e. be independent of system-specific d15N variability).  We investigated these criteria in invertebrate communities in 16 sites along a gradient from mountain to prairie streams which also represent a gradient in land use.

@@

We sampled sixteen sites in three streams that span the Rocky Mountain to Great Plains ecotone (Figure 1).  We selected sites  to represent an environmental gradient based on elevation and land use that resulted in a wide variation of d15N values at the base of local food webs. All streams are snowmelt dominated with peak flows occurring in the spring and base flows occurring in mid-summer. Their watersheds transition from forested land at high elevations to agriculturally dominated land with some urbanization at lower elevations.

### Longitudinal position index

We used a synthetic variable developed by Maitland (2020) to index a site’s relative location along the Rocky Mountain--Great Plains gradient. Briefly, The variable combines seven metrics of environmental conditions at a site into an index of longitudinal stream position using the Axis 1 score from a principal component analysis (PCA). Input variables to the PCA were elevation (m a.s.l.), stream slope (km/km), distance to North Platte River (km), Strahler stream order, mean August water temperature (℃), upstream drainage area (km2), and stream width (m). The synthetic index (PC1) explained 78% of site-level variation among the 16 study sites and described changes in environmental conditions from upstream, high-elevation, cold-water streams to downstream, low-elevation, warm-water streams.

The synthetic index is positively correlated with d15N values of basal resources and primary consumers (Maitland 2020). d15N values increase moving from upstream to downstream, likely due to changing land use. δ15N of aquatic biota can be highly correlated with nitrogen (% of total N load) from livestock and fertilizers (Anderson and Cabana 2006), and in this system, the relative amount of natural land cover decreases and pastureland and agricultural land increase moving downstream. Therefore, we used the synthetic index (PC1) a priori as the environmental variable determining background δ15N levels.

### Sample collection

Samples were collected during summer of 2016. Detailed sampling proceedure are in Maitland (2020). Breifely, basal resources were sampled  in June, July, and August to account for fine-scale temporal variation in d15N (Jardine et al., 2014).  Seston was collected by filtering three replicates of up to 10 L of water onto pre-combusted (550 °C, 4 hours) filters (Whatman GF/F, 47 mm O) in the field using a modified portable drill pump (Kelso & Baker, 2016). Biofilm was collected by scraping five cobble-sized rocks within the reach.  invertebrate were collected during the same three sampling events during the summer of 2016. A D-framed kick net was used in a variety of habitats (i.e. riffle, pools, macrophytes, and river margins) to collect a representative sample of the macroinvertebrate assemblage.  invertebrate were kept in filtered stream water for 24 hours to allow for gut clearance, then frozen. Muscle tissue samples were obtained from fish collected during the August sampling event using a backpack electrofishing unit (Smith-Root, Vancouver, WA).  Brown Trout Salmo trutta (Linnaeus, 1758), Creek Chub Semotilus atromaculatus (Person, Year), White Sucker Catostomus commersonii (Lacépède, 1803), were used because these species were found at greater than 50% of the sites.  To minimize the number of fish euthanized, a muscle plug using a 5-mm biopsy punch was taken from fish greater than 300 mm in length. The remaining fish used for SIA were euthanized with a lethal dose of MS-222, placed on ice and then frozen upon returning to the lab. Those fish not selected for SIA were returned to the stream.

invertebrate were identified to family level using an invertebrate guide (Merritt et al. 2008). We classified invertebrate into FFGs using a trait-based method. We obtained invertebrate trophic trait data from the USA Freshwater Biological Traits Database (Vieira et al. 2006), and then assigned trait affinity scores from 0 (absent) to 3 (strong affinity) using “fuzzy-coding” (Chevene et al. 1994). Because larvae and adults have different feeding behaviors, Elmidae were separated by life stage during identification.

@@

<!-- \begin{equation} -->
<!-- \delta^{n} = (\frac{R_{sample}}{R_{standard}}+1) * 100 (\#eq:ratio) -->
<!-- \end{equation} -->

<!-- where dN is either N15 or c13, and R is the ratio of the heavy to lighter isotope. Analytical error (i.e., 1 SD of lab standard) of sample runs was estimated at 0.07‰ (d13C) and 0.18‰ (d15N) for the invertebrates, and 0.04‰ (d13C) and 0.10‰ (d15N) for the fish.  -->

### Baseline criteria and data analysis

We used the four criteria proposed by Kristensen et al. (2016) to assess the suitability of invertebrate taxa for use as isotopic baselines.  The criteria are: (1) have a wide geographic distribution, (2) have low with-in site variation of d15N values, indicating a uniform nonomnivorous feeding mode, (3) have d15N values that correlate the longitudinal gradient index, indicating they reflect background d15N variability, (4) provide trophic position estimates of secondary consumers independent of systematic d15N variability.

@@

To assess the geographic distribution of different taxa and FFGs, we calculated the percent of sites where the taxon or group was present.

@@

To assess within-site d15 variability among groups, we calculated the coefficient of variation (CV) for d15N values as SD/mean. A low coefficient of variation for d15N indicates that a baseline candidate has a non-omnivorous diet.

We used one-way analysis of variance (ANOVA) to test for differences in within-site variation (CV) among baselines. Separate analyses were done on individual taxa and FFGs.  The CV values were square root or log transformed to ensure normality.  We used the Tukey HSD post hoc tests for pairwise-comparison using an alpha of 0.05.

@@

For all invertebrate taxa and FFGs, we used least squares linear regression to assess relationships between the baseline’s d15N position at sites along the longitudinal stream gradient.  We wished to compare our baselines’ relationships to the longitudinal gradient with those of the basal resources. So, we did separate least squares linear regressions for two of the basal resources at our sites and compared R2 values between the baselines and the resources at our sites.  The R2 value of our basal resources represent the minimum level of variation that our alternate baselines should explain in order to be used as alternatives to the traditional method (e.g. all basal resources across a season).   
<!-- Baselines with significant ($P \le 0.05$) relationships with the longitudinal gradient and that had R2 values equal to or greater than the basal resources were considered suitable baselines. -->
